---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

# Reproducible Research - Peer Assessment 1 <br><br>


## Loading and preprocessing the data  

```{r, echo = TRUE}
unzip("activity.zip")
df1 <- read.csv("activity.csv")
df1$date <- as.Date(df1$date, "%Y-%m-%d")
```
<br><br>

## What is mean total number of steps taken per day?  

```{r TotalSteps, echo = TRUE}
tSteps <- with(df1, tapply(steps, date, sum, na.rm=TRUE))
hist(tSteps, 15, main="", xlab="Total Steps per Day", ylim=c(0,20))
meanSteps <- mean(tSteps, na.rm=TRUE)
medSteps <- median(tSteps, na.rm=TRUE)
print(paste("The mean total number of steps per day is", meanSteps))
print(paste("The median total number of steps per day is", medSteps))
```
<br><br>

## What is the average daily activity pattern?  

```{r TimeSeries, echo = TRUE}
avgD <- with(df1, tapply(steps, interval, mean, na.rm=TRUE))
intervals <- unique(df1$interval)
df2 <- data.frame(intervals, avgD)
with (df2, plot(intervals, avgD, type="l", xlab="Interval", ylab="Average Number of Steps"))
maxInt <- df2[which.max(df2$avgD),"intervals"]
print(paste("The interval containing the maximum number of steps is", maxInt))
```
<br><br>

## Imputing missing values  

```{r, echo = TRUE}
mValues <- with(df1, sum(is.na(steps)))
print(paste("The total number of missing values is", mValues))
```


The mean of each 5-min interval was rounded and substituted in place of the missing values  
```{r MissingValues, echo = TRUE}
library(data.table)
dfNew <- copy(df1)
mIndex <- which(is.na(df1$steps))
for (i in mIndex){
	missInt <- df1$interval[i]
	dfNew$steps[i] <- round(df2[df2$intervals==missInt, "avgD"])
}
newSteps <- with(dfNew, tapply(steps, date, sum))

hist(newSteps, 15, main="", xlab="Total Steps per Day", ylim=c(0,25))
print(paste("The mean total number of steps per day is", mean(newSteps)))
print(paste("The median total number of steps per day is", median(newSteps)))
```
  
The impact of imputing missing values was that there was a modest increase in the mean and median compared to the original data. The median increased from 10395 to 10762 (3.5% increase). The mean increased from about 9354 to 10765 (15% increase). 

<br><br>

## Are there differences in activity patterns between weekdays and weekends?  


*Create dataset with new factor variable with two levels: "weekday" and "weekend"*
```{r, echo=TRUE}
timeInt <- as.integer(levels(as.factor(dfNew$interval)))
timeInt <- format(timeInt, format="%H%M")
timeInt <- gsub(" ", "0", timeInt)
w <- weekdays(as.POSIXct(paste(dfNew$date, timeInt)))
wkday <- c("Monday",   "Tuesday",   "Wednesday", "Thursday",  "Friday")
wkend <- c("Saturday", "Sunday")
for (i in 1:length(w)){
	if (w[i] %in% wkday) { dfNew$day[i] <- "weekday" } 
		else if (w[i] %in% wkend) { dfNew$day[i] <- "weekend" }
}
dfNew$day <- as.factor(dfNew$day)
```
<br><br>
*Create panel plot to compare weekday and weekend activities*  

```{r TimeSeriesPanel, echo=TRUE}
sp <- split(dfNew, dfNew$day)
wkdayData <- with(sp[["weekday"]], tapply(steps, interval, mean))
wkendData <- with(sp[["weekend"]], tapply(steps, interval, mean))

rng<-range(wkdayData)
par(mar=c(4, 4, 2, 2) + 0.1)
par(mfrow=c(2,1))
plot(intervals, wkdayData, type="l", xlab="", ylab="Steps", main="Weekday", ylim=rng)
plot(intervals, wkendData, type="l", xlab="Interval", ylab="Steps", main="Weekend", ylim=rng)
```
